---
title: "Lung Cancer Data || Kaplan-Meier Curve || R"
description: |
  The Kaplan-Meier curve visualizes survival probabilities over time, helping to identify survival trends and the impact of variables like sex on outcomes, crucial for understanding lung cancer progression.
author:
  - name: Prasanth Kumar
    url: https://www.linkedin.com/in/prasanthkumar28/
date: 08-17-2024
output:
  distill::distill_article:
    self_contained: false
    df_print: kable
---

## Setup Chunks in RMarkdown

```{r Chunk Setup, include=FALSE}
knitr::opts_chunk$set(comment = "", message = FALSE, warning =FALSE,
                      echo = FALSE)
```


## Loading Required Libraries

 Loading several essential libraries, including "tidyverse". These libraries offer diverse functionalities for data manipulation, visualization, and reporting.

 And also we need the "survival" package for the NCCTG Lung Cancer data set in it, for graphical representation "survminer" package is used.
 
```{r Library}
library(tidyverse)
library(survival)
library(survminer)
library(knitr)
```


## Data Description 

### NCCTG Lung Cancer Data

Survival in patients with advanced lung cancer from the North Central Cancer Treatment Group. Performance scores rate how well the patient can perform usual daily activities.


### Format 

inst:	Institution code.  
time:	Survival time in days.  
status:	censoring status 1=censored, 2=dead.  
age:	Age in years.  
sex:	Male=1 Female=2.  
ph.ecog:	ECOG performance score as rated by the physician. 0=asymptomatic, 1= symptomatic but completely ambulatory, 2= in bed <50% of the day, 3= in bed > 50% of the day but not bedbound, 4 = bedbound.  
ph.karno:	Karnofsky performance score (bad=0-good=100) rated by physician.    
pat.karno:	Karnofsky performance score as rated by patient.    
meal.cal:	Calories consumed at meals.  
wt.loss:	Weight loss in last six months (pounds).  

```{r Lung Dataset}
lung_data <- lung 
kable(head(lung_data))
```


## Data Cleaning 

```{r Data Cleaning}
lung_data <- na.omit(lung_data)

colSums(is.na(lung_data))
```

## Survival Analysis

## Kaplan-Meier Curve
  
  In this analysis, we use the Kaplan-Meier curve to visualize the overall survival probability of patients in the dataset. It provides a visual representation of survival probability over time, where each step down in the curve indicates an event(e.g., death).
```{r KM Curve}
# Survival Object
surv_obj <- Surv(time = lung_data$time, event = lung_data$status)

# Fit the Kaplan_Meier Model
fit <- survfit(surv_obj ~ 1, data = lung_data)

# Plot the Kaplan - Meier Curve 
ggsurvplot(fit, data = lung_data,
           xlab = "Time in Days",
           ylab = "Surivial Probability",
           title = "Kaplan-Meier Survival Curve",
           conf.int = T, # Confidence Interval
           ggtheme = theme_minimal()) # Minimal Theme
```

## Kaplan-Meier Curve by Sex
 We stratify the Kaplan-Meier Curve by the 'sex' variable, dividing the data into male and female patients. While the overall KM Curve is informative, survival probabilities can vary significantly between different subgroups within the data. 
 
 Such comparisons are crucial for understanding the impact of demographic or clinical variables on Survival outcomes.
 
```{r KM Curve - Sex}
fit_s <- survfit(surv_obj ~ sex, data = lung_data)

# Plot the Kaplan - Meier Curve 
ggsurvplot(fit_s, data = lung_data,
           xlab = "Time in Days",
           ylab = "Surivial Probability",
           title = "Kaplan-Meier Survival Curve by Sex",
           conf.int = T, # Confidence Interval
           ggtheme = theme_minimal(), # Minimal Theme
           palette = c("palegreen", "pink"),# Set colors for the curvers
           legend.labs = c("Male", "Female")) # Label the Legend 
```

